---
title: "Day_03_scientific_visualization_training"
author: "By Muhammad Yasir Qurashi"
date: "2026-03-03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Two-way ANOVA in R**

## Loading libraries

Install these packages before activation if you they are not pre-installed in your environment

```{r}
library(ggplot2)
library(multcompView)
library(ggpubr)
library(tidyverse)
library(ggrain)
library(ggsci)
library(RColorBrewer)
```

## Creating/loading dataset

```{r}
TG <- ToothGrowth
TG$dose <- as.factor(TG$dose)
summary(TG)
```

# **Distribution of Data**

## Raincloud plot

```{r, fig.width=7, fig.height=4}
p1 <- ggplot(data = TG, mapping = aes(x = dose, y = len, fill = dose)) +
  geom_rain(color = "blue")+
  labs(x = "Dose", y = "Tooth Length (mm)", title = "Rain Cloud Plot", subtitle = "Distribution of tooth length by dose", caption = "Made by Muhammad Yasir Qurashi")+
  facet_wrap(~supp)+
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 16, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    legend.background = element_rect(color = "black", fill = "grey"),
    legend.position = "right",
    legend.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "grey", width = 3),
    strip.text = element_text(family = "bold"),
    panel.border = element_rect(linewidth = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  )+
  scale_fill_bmj();p1
```

## Density Plot

```{r fig.height=4, fig.width=7}
p2 <- ggplot(TG, aes(x = len, fill = dose)) +
  geom_density(alpha = 0.6) +
  labs(
    title = "Density Plot",
    x = "Tooth Length",
    y = "Density", subtitle = "Distribution of tooth length by dose", caption = "Made by Muhammad Yasir Qurashi")+
  facet_wrap(~supp)+
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 16, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    legend.background = element_rect(color = "black", fill = "grey"),
    legend.position = "right",
    legend.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "grey", width = 3),
    strip.text = element_text(family = "bold"),
    panel.border = element_rect(linewidth = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  )+
  scale_fill_brewer(palette = "Set1", direction = 1);p2
```

## Hypothesis testing steps for one-way ANOVA

**1. Make your hypothesis**

null hypothesis --\> there is no significant interactive effect of dose & supp on tooth length

alternative --\>there is significant interactive effect of dose & supp on tooth length

**2. Define your significance level**

alpha = 0.05, C.I = 0.95

**3. Selection of test**

*Two way ANOVA*

**4. Calculation**

```{r}
# Normality Check ( first condition of ANOVA)

TG %>% 
  group_by(dose, supp) %>% 
  summarise(
  shapiro_p = shapiro.test(len)$p.value
  )

 # Data is normal
```

```{r}
# Homogenity ( Second Condition for ANOVA)
library(car)
leveneTest(len ~ dose*supp, data = TG, alternative = "two.sided")
 # Data is Homogenous too
```

```{r}
# ANOVA calculation
results_anova <- aov(len ~ dose*supp, data = TG)
summary(results_anova)
```

**5. Conclusion**

Two way ANOVA has been done with statistical significant results \* ( n = 60, p_value \< 0.05 )

# **Graphical Presentation**

## Bargraph without lettering

```{r, fig.width=8, fig.height=4}
p3 <- ggplot(data = TG, mapping = aes(x = dose, y = len, fill = dose))+
  geom_bar(stat = "summary", fun = "mean", width = 0.3, linewidth = 0.5)+
  geom_errorbar(stat = "summary", fun.data = mean_se, width = 0.05, linewidth = 0.7)+
  labs(x = "Dose", y = "Mean Tooth Length (mm)", title = "Bar Plot", subtitle = "Two-way ANOVA without lettering", caption = "made by Muhammad Yasir Qurashi")+
  stat_compare_means(method = "anova", label.y = 26)+
  facet_wrap(~supp)+
    theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 16, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    legend.background = element_rect(color = "black", fill = "grey"),
    legend.position = "right",
    legend.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "grey", width = 3),
    strip.text = element_text(family = "bold"),
    panel.border = element_rect(linewidth = 1),
     panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  ) +
  scale_fill_brewer(palette = "Set2", direction = 1);p3
```


## Barplot with lettering

TO make barplot with significant letters there are four steps given below:

**1. Step 1**

Find group wise mean

```{r}
dose_mean <- TG %>% 
  group_by(dose, supp) %>% 
  summarise(
    mean_len = mean(len),
    sd = sd(len)
  );dose_mean
```

**1. Step 2**

calculation of anova

```{r}
anova <- aov(len ~ dose*supp, data = TG)
summary(anova)
```

**1. Step 3**

Post_hoc test

```{r}
tukey <- TukeyHSD(anova)
print(tukey)
```


**1. Step 4**

Giving Significant Letters

```{r}
SL <-  multcompLetters4(anova,tukey)
print(SL)
SL <- data.frame(SL$`dose:supp`$Letters)
SL
```


**1. Step 5**

Combine Significant letters with Mean Table

```{r}
dose_mean <- cbind(dose_mean, SL = SL$SL..dose.supp..Letters)
dose_mean
```

## Plot

```{r, fig.width=10, fig.height=4}
p4 <- ggplot(data = TG, mapping = aes(x = dose, y = len, fill = dose))+
  geom_bar(stat = "summary", fun = "mean", width = 0.5, linewidth = 0.5)+
  geom_errorbar(stat = "summary", fun.data = mean_se, width = 0.05, linewidth = 0.7)+
  geom_text(data = dose_mean, aes(x = dose, y = mean_len, label = SL), size = 5, hjust = 0, vjust = -1)+
  facet_wrap(~supp)+
  labs(x = "Dose", y = "Tooth Length Mean (mm)", title = "Bar Plot", subtitle = "Two-way ANOVA with lettering", caption = "By Muhammad Yasir Qurashi" )+
    theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 16, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    legend.background = element_rect(color = "black", fill = "grey"),
    legend.position = "right",
    legend.text = element_text(size = 10, face = "bold"),
    strip.background = element_rect(color = "black", fill = "grey", width = 3),
    strip.text = element_text(family = "bold"),
    panel.border = element_rect(linewidth = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  ) +
  scale_fill_brewer(palette = "Dark2", direction = 1)+
  ylim(0,30);p4
```

## Combine Plot

```{r, fig.height = 8, fig.width = 14}
library(patchwork)

finalplot <- (p1 | p2 ) / (p3 | p4) +
  plot_annotation(tag_levels = "A");finalplot
```


Best Regards,

*Muhammad Yasir Qurashi*

Research Data Analysis Tools Mentor