---
title: "Day_04_scientific_visualization_training"
author: "By Muhammad Yasir Qurashi"
date: "2026-03-06"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Correlation & Multivariate Relationship Visualization**

## Loading libraries

```{r}
library(ggplot2)
library(multcompView)
library(ggpubr)
library(tidyverse)
library(ggrain)
library(ggsci)
library(RColorBrewer)
```

## Loading dataset

```{r}
mt_data <- mtcars
summary(mt_data)
class(mt_data)
```

## Variables Interpretation

| Variable | Meaning                            |
|----------|------------------------------------|
| mpg      | Miles per gallon (fuel efficiency) |
| wt       | Car weight (1000 lbs)              |
| hp       | Gross horsepower                   |
| disp     | Engine displacement                |
| cyl      | Number of cylinders                |

## Scientific Research Questions to be Addressed

Is vehicle weight negatively correlated with fuel efficiency (mile per galloon) ?

Does horsepower predict fuel consumption ?

Are engine size, horsepower, and weight inter-correlated ?

## Scatterplot

```{r}
p1 <- ggplot(mt_data, mapping = aes(x = wt, y = mpg)) +
  geom_point(size = 2, color = "blue", alpha = 1) +
  labs(x = "Vehicle weight", y = "Fuel efficiency (mpg)", title = "Scatter Plot", subtitle = "Vehicle Weight Vs mpg", caption = "Made by Muhammad Yasir Qurashi") +
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 16, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  );p1
```

## Correlation Plot

```{r}
p2 <- ggplot(mt_data, mapping = aes(x = wt, y = mpg)) +
  geom_point(size = 2, color = "blue", alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black")+
  stat_cor(method = "pearson", label.x = 3, label.y = 30)+
  stat_regline_equation(label.x = 3, label.y = 28)+
  labs(x = "Vehicle weight", y = "Fuel efficiency (mpg)", title = "Scatter Plot", subtitle = "Correlation of wt Vs mpg", caption = "Made by Muhammad Yasir Qurashi") +
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 20, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  );p2
```

## Correlation Plot with 3 variables

```{r, fig.height=4, fig.width=8}
p3 <- ggplot(mt_data, mapping = aes(x = wt, y = mpg, colour = factor(cyl))) +
  geom_point(size = 3, alpha = 1) +
  geom_smooth(method = "lm", se = F, width = 2)+
  stat_regline_equation(label.x = c(2,3,4))+
  facet_wrap(~cyl)+
  labs(color = "Cylinders", x = "Weight", y = "Fuel efficiency", title = "Scatter Plot", subtitle = "Correlation by No of Cylinders", caption = "Made by Muhammad Yasir Qurashi") + theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 20, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.background = element_rect(color = "black", fill = "lightgrey", linewidth = 1),
    strip.text = element_text(color = "black", face = "bold"),
    panel.border = element_rect(color = "black", linewidth = 1),
    legend.background = element_rect(fill = "lightgrey", color = "black")
  ) +
  scale_color_brewer(palette = "Dark2");p3
```

```{r, fig.height=7, fig.width=12}
library(GGally)
p4 <- ggpairs(
  mtcars,
  columns = c("mpg", "wt", "hp", "disp"),
  aes(color = factor(cyl), alpha = 0.7),
  lower = list(continuous = wrap("points", alpha = 0.6, size = 2)),
  upper = list(continuous = wrap("cor", size = 5)),
  diag = list(continuous = wrap("densityDiag"))
) +
  scale_fill_brewer(palette = "Dark2", direction = 1)+
  labs(color = "Cylinders",caption = "Made by Muhammad Yasir Qurashi") + 
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 20, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    strip.background = element_rect(color = "black", fill = "lightgrey", linewidth = 1),
    strip.text = element_text(color = "black", face = "bold"),
    panel.border = element_rect(color = "black", linewidth = 1));p4
```

## What Happens here

ggpairs(), automatic creates

| Panel          | Visualization                    |
|----------------|----------------------------------|
| Diagonal       | Distribution (histogram/density) |
| Lower triangle | Scatterplots                     |
| Upper triangle | Correlation coefficients         |


## Interpretation

“Vehicle weight was strongly negatively correlated with fuel efficiency (r = -0.87, p < 0.001). Linear regression analysis indicated that heavier vehicles exhibited significantly reduced miles per gallon. Engine displacement and horsepower were positively correlated (r = 0.79), suggesting structural coupling between engine size and power output.”

# Compile figure


```{r, fig.height=8, fig.width=12}
library(patchwork)

final_fig <- (p1 | p2) / (p3) +
  plot_annotation(tag_levels = "A");final_fig

```


Best Regards,

*Muhammad Yasir Qurashi*

Research Data Analysis Tools Mentor
