---
title: "Day_05_scientific_visualization_training"
author: "By Muhammad Yasir Qurashi"
date: "2026-03-08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Principal Component analysis in R**

Principal Component Analysis is a dimensionality reduction technique that transforms correlated variables into a smaller number of uncorrelated components that capture the major variation in the dataset

## Scientific Questions

* Which states have similar crime patterns?

* Which crime types drive variation across states?

* Can states be clustered into crime profiles?

## Loading dataset

```{r}
UA <- USArrests
UA
summary(UA)
head(UA)
tail(UA)
```

## Loading Required packages

```{r}
library(factoextra)
library(cluster)
library(ggfortify)
library(tidyverse)
```

# **Principal Component Analysis (PCA)**

## Step:01 Data Standardization

Why we do it

Variables have different scales, if we do not standardize them variable with large scale dominates the other variable

```{r}
scale_UA <- scale(UA)

# What does scale mean

# mean = 0
# standard deviation = 1
# Now each variable contributes equally.

```

## Step 02 Run PCA

Why we do it

PCA identifies major patterns of variation

```{r}
pca_result <- prcomp(scale_UA)
pca_result
summary(pca_result)
```

## Step 03 Scree plot

Why we use it

The scree plot answers, how many principal components should we keep?

It shows variance explained by each PC.

```{r, fig.height=5, fig.width=6}


p1 <- fviz_eig(pca_result, addlabels = T, ylim = c(0, 65), barfill = "lightblue", barcolor = "black") +
  theme_classic() +
  labs( title = "Scree plot", x = "Dimensions", y = "Percentage of explained variance", caption = "Made by Muhammad Yasir Qurashi") +
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    plot.title = element_text(size = 16, color = "black", face = "bold", hjust = 0.5)
  );p1

# Iterpretation

# Pca1 & pca2 shows 87% variation, since most information in dataset is captured by first two components
```

## Step 04 Variables correlation circle

Why this is powerful

This plot visualizes relationships among variables.

It shows:

-   variable correlations

-   variable influence

-   variable grouping

```{r, fig.height=5, fig.width=6}
p2 <- fviz_pca_var( pca_result, col.var = "cos2", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), repel = TRUE ) +
  theme_classic() +
  labs( title = "PCA Of parameters", x = "PC1 (62%)", y = "PC2 (24.7)", caption = "Made by Muhammad Yasir Qurashi") +
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    plot.title = element_text(size = 16, color = "black", face = "bold", hjust = 0.5)
  );p2
```

*Interpretation rules*

What the Circle Means

* The circle represents maximum correlation (r = 1)

| Pattern               | Meaning                     |
| --------------------- | --------------------------- |
| Arrows close together | strong positive correlation |
| Opposite arrows       | negative correlation        |
| 90° angle             | no correlation              |
| Long arrow            | strong variable influence   |

In above figure; Murder and assault variables clustered closely together in the correlation circle, indicating strong positive correlation between these crime indicators across states.

## Step 05 Checking Contribution of each variable to PCA components

```{r, fig.height=8}
p4 <- fviz_contrib(pca_result, choice = "var", axes = 1, label = T); fviz_contrib(pca_result, choice = "var", axes = 2, label = T);  fviz_contrib(pca_result, choice = "var", axes = 3, label = T); fviz_contrib(pca_result, choice = "var", axes = 4, label = T);p4
```

## K_means Clustering of PCA scores

```{r, fig.height=4, fig.width=5}

k_cluster <- kmeans(scale_UA, centers = 4)
k_cluster

scores <- as.data.frame(pca_result$x)

head(scores)

scores$Cluster <- factor(k_cluster$cluster)

p3 <- ggplot(scores, aes(PC1, PC2, color = Cluster, fill = Cluster)) +
  stat_ellipse( geom = "polygon", alpha = 0.3, color = "black") +
  geom_point( shape = 21, size = 3, color = "black" ) +
  theme_classic() +
   labs( title = "Clustering of PCA components", x = "PC1 (62%)", y = "PC2 (24.7)", caption = "Made by Muhammad Yasir Qurashi") +
  theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    plot.title = element_text(size = 16, color = "black", face = "bold", hjust = 0.5)
  );p3
```

Best Regards,

*Muhammad Yasir Qurashi*

Research Data Analysis Tools Mentor
